The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting

### 2024-12-13 manuscript/main.tex
final draft for submission

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{todonotes,mathtools,bm,amsmath,mathpazo}
\mathtoolsset{showonlyrefs}
\setlength{\parindent}{0cm}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\journal{Journal of}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Pharmaceutical Supply Chain Forecasting by using Machine Learning in Ethiopian Pharmaceutical Supply Services},
  pdfauthor={Arebu Issa Bilal; Arebu Issa Bilal; Bahman Rostami-Tabar; Umit Sezer Bititci; Teferi Gedif Fenta},
  pdfkeywords={Pharmaceutical Demand Forecasting, Ethiopian
Pharmaceutical Supply Services, ARIMA Model, Forecasting
Variables, Forecast Accuracy},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Pharmaceutical Supply Chain Forecasting by using Machine Learning
in Ethiopian Pharmaceutical Supply Services}
\author[1]{Arebu Issa Bilal%
\corref{cor1}%
}
 \ead{arebu.issa@aau.edu.et} 
\author[]{Arebu Issa Bilal%
%
}
 \ead{arebu.issa@aau.edu.et} 
\author[2]{Bahman Rostami-Tabar%
%
}
 \ead{rostami-tabarb@cardiff.ac.uk} 
\author[3]{Umit Sezer Bititci%
%
}
 \ead{U.S.Bititci@hw.ac.uk} 
\author[1]{Teferi Gedif Fenta%
%
}
 \ead{tgedif@gmail.com}

\affiliation[1]{organization={Addis Ababa University, Department of
Pharmaceutics and Social Pharmacy at School of
Pharmacy},country={Ethiopia},countrysep={,},postcode={9086},postcodesep={}}
\affiliation[2]{organization={University, Cardiff Business
School},country={UK},countrysep={,},postcode={CF5},postcodesep={}}
\affiliation[3]{organization={University, Edinburgh Business
School},country={UK},countrysep={,},postcode={EH14 4AS},postcodesep={}}

@@

\begin{abstract}
This study delves into the realm of pharmaceutical demand forecasting
within the Ethiopian Pharmaceuticals Supply Services(EPSS) context. It
encompasses a comprehensive analysis of 37 key pharmaceutical
products,assessing their forecast accuracy across variation models. The
initial assessment reveals that the Autoregressive Integrated Moving
Average (ARIMA) model emerges as a frotrunner, boasting a Root Mean
Squared Error (RMSSE) value of 0.850. ARIMA's prowess in capturing
complex demand patterns suggests its suitability for pharmaceutical
sales forecasting in the EPSS landscape. To argument forecasting
accuarcy, predictor varaibles such as physical count, budget release and
closure by the goverment, the impact of COVID-19, and promotional
campaigns were incorpoarted. Remarkably, the ARIMA model mantainsits
superiority even in the presences of these external factors, reiterating
its roustness. Delving further into individual pharmaceutical products,
ARIMA with predictors demonstrates exceptional performance for a
substantial portion (32.43\%) of the products, including Amoxicillin
500mg capsule, Anti RhD, Artehmether + Lumefantrine, and others.
Conversely,alternative models like naive, mean, and regression methods
showcase strengths for different pharmaceutical categories, underscoring
the importance of tailoring forecasting approaches to product-specific
dynamics. While this study significantly advances the discourse on
pharmaceutical demand forecasting, limitations stemming from data
quality and external variables underscore the intricacies of the
forecasting landscape. In conclusion, it advocates the integration of
the ARIMA model with predictor variables as a strategic approach for
refining pharmaceutical demand forecasting within EPSS. It also
encourages further exploration of advanced forecasting techniques and
ongoing refinement to bolster supply chain efficiency and ensure
consistent pharmaceutical availability. Capacity building through staff
training and replication of the study in diverse healthcare settings are
recommended steps toward resilient and efficient pharmaceutical supply
chains.These findings hold promise for fostering supply chains that are
adaptable, responsive, and capable of ensuring the continuous
availability of essential pharmaceuticals in Ethiopia.
\end{abstract}

\begin{keyword}
    Pharmaceutical Demand Forecasting \sep Ethiopian Pharmaceutical
Supply Services \sep ARIMA Model \sep Forecasting Variables \sep 
    Forecast Accuracy
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, enhanced, frame hidden, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi

\hypertarget{sec-intro}{%
\section{Introduction}\label{sec-intro}}

Universal access essential medicines is a cornerstone of effective
healthcare systems, yet ensuring equitable availability of these vital
medications remains an urgent global
challenge\citep{quick2003essential, world2004annual}. Across continents,
especially in Africa and Asia, many nations grapple with restricted
access due to factors like unaffordability and suboptimal supply chain
management \citep{world2004medicines}. A survey conducted across eight
sub-Saharan African countries indicated that there were unacceptably low
availability of essential medicines.The mean availability of 12 priority
essential medicines for women ranged from 22\% to 40\% and from 28\% to
57\% for children \citep{droti2019poor}. However, the issue of drug
shortages extends beyond developing countries, with instances reported
in the United States and Europe, adversely impacting healthcare systems
and compromising patient care quality
\citetext{\citealp[\citet{kaakeh2011impact}]{fox2003managing}; \citealp{johnson2011drug}; \citealp{huys2013european}; \citealp{le2011prevalence}}.

The consequences of drug shortages are felt both financially within the
healthcare sector and in terms of the quality of patient care
\citep{alspach2012drug, kaakeh2011impact, kaakeh2011impact, baumer2004national}.The
consequences extend across patients, providers, healthcare institutions,
and research programs. Patients face delays or cancellations of
treatments, substandard care due to the unavailability of optimal
therapies, and increased costs in a secondary ``grey market.''
Compromised patient safety, complications, prolonged hospital stays, and
even deaths have been reported {[}\citet{alspach2012drug};{]}.Addressing
these shortages is crucial to preserving the quality of patient care and
ensuring the effectiveness of healthcare delivery. Thus, a smoothly
operating pharmaceutical supply chain is essential to ensure timely
access to essential medicines, making accurate demand forecasting an
indispensable component.

An efficient pharmaceutical supply chain is vital for timely access to
essential health commodities. One of the most crucial components of such
a supply chain is a well-defined demand forecasting process. Forecasting
accurately the future demand for essential medicines provides valuable
insights for anticipating future healthcare requirements including
procurement and replenishment of
inventory\citep{subramanian2021effective}. Despite the significance of
forecasting for supply chain management, pharmaceutical product
forecasting remains a complex endeavor influenced by a multitude of
factors. Key considerations include the availability of data, the
specific position within the supply chain, and the nuanced understanding
of contributing elements. The position within the supply chain
introduces unique challenges and dynamics, requiring tailored
forecasting approaches. Additionally, a nuanced understanding of
contributing elements involves grasping intricate factors that impact
forecasting accuracy, such as market trends, regulatory changes, and
external influences. Schneider, Wilson, and Rosenbeck (2010) and Hyndman
and Athanasopoulos (2018) highlight the intricate nature of
pharmaceutical forecasting, emphasizing the need for a comprehensive
approach that incorporates these multifaceted aspects to enhance supply
chain management
effectiveness\citep{schneider2010pharmaceutical, hyndman2018forecasting}.

The Ethiopian pharmaceutical sector's distinct characteristics, marked
by a diverse array of medicines and unique data challenges stemming from
service delivery points, staff capacity limitations, communication
hurdles, prolonged procurement and distribution lead times, poor
forecast accuracy, and policy intricacies, necessitates tailored
forecasting strategies that cater to the nation's specific context
\citep{boche2022procurement}. In this context, a comprehensive
exploration of effective forecasting models that align with Ethiopia's
pharmaceutical landscape is paramount to mitigating stock outs and
enhancing healthcare delivery. The implications of pharmaceutical demand
forecasting extend beyond financial considerations to societal
well-being \citep{rostami2022forecasting}. In Ethiopia, a country
undergoing demographic and healthcare evolution, precise demand
forecasting assumes immense importance for resource allocation, supply
chain optimization, and policy formulation. With mounting interest in
refining demand forecasting in Ethiopia's pharmaceutical sector, this
study seeks to contribute to existing knowledge by investigating and
recommending proficient demand forecasting methods. Our investigation
centers on the application of mean, naive, exponential smoothing,
regression, and ARIMA models to bolster forecast accuracy and
effectiveness. This paper aims to forecast the demand for selected
pharmaceuticals in EPSS over the next 12 months by utilizing various
forecasting models and integrating predictors. The integration of
predictors with the models will be examined to determine whether they
enhance the forecasting accuracy. These predictors are derived from a
historical analysis of events in Ethiopia over the last 5 years that
have impacted pharmaceutical sales in EPSS. These events include
COVID-19, conflicts in the country, disease campaigns, budget closures,
budget releases, and physical inventory, all of which, in varying
degrees, influence the demand for pharmaceuticals in Ethiopia.

The current paper offers the following contributions we aim at
forecasting pharmaceutical sales over a 12-month period these models aim
to enhance the accuracy and reliability of demand predictions within the
context of EPSS. Moreover an essential contribution lies in the
identification of key factors influencing pharmaceutical sales. The
study develops a model that effectively incorporates these driving
factors, providing a more nuanced and comprehensive approach to demand
prediction. We generate forecasts and evaluate model performance using
the Root Mean Squared Error (RMSE), Root Mean Squared Scaled Error
(RMSSE) and Mean Absolute Error (MAE), Mean Absolute Scaled Error
(MASE), Crips and Winkler are used to assess probabilistic forecast
accuracy. \#how you know the forecasting accutacy problem and provide
eveidence about the forecasting problems \# most forecast focus on point
forecast and we want to focus on probablistic forecast \# i will write
the contribution of the reearch rather than these objectives

This research is guided by several objectives:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Evaluate the robustness of various forecasting models for
  pharmaceutical demand prediction within EPSS.
\item
  Identify pharmaceuticals that exhibit superior forecast accuracy under
  specific models.
\item
  Assess the performance of forecast accuracy metrics to gauge the
  efficacy of selected models.
\item
  Test the incorporation of diverse predictors affecting forecasts and
  evaluate resulting forecasting accuracy. In pursuing these objectives,
  this study endeavors to elevate the precision and reliability of
  pharmaceutical demand forecasting, thereby contributing to optimized
  supply chain management and enhanced healthcare outcomes in Ethiopia.
\end{enumerate}

\hypertarget{sec-lit}{%
\section{Research background}\label{sec-lit}}

Precise sales prediction is an essential and inexpensive way for each
company to augment their profits, decrease their costs, and achieve
greater flexibility to changes. In other words, exact sales forecasting
is utilized for capturing the tradeoff between customer demand
satisfaction and inventory costs \citep{gupta2000mid}. Especially, for
the pharmaceutical industry, successful sales forecasting systems can be
very beneficial, due to the short shelf-life of many pharmaceutical
products and the importance of the product quality which is closely
related to the human health \citep{doganis2006time}.There are many
decisions that depend on the quality of forecasts in the health care
system, from capacity planning to layout decisions to the daily
schedules. In general,the role of forecasting in health care is to
inform both clinical and non-clinical decisions. While the former
concerns decisions related to patients and their treatments
\citep{makridakis2019forecasting}, the latter involves
policy/management, and supply chain decisions that support the delivery
of high-quality care for patients.In the coming paragarphs we will
review what other researchers had done in the area of pharmaceuticals
forecasting.Morover today,manufacturing companies are organized
following the principles of just-in-time (JIT) delivery which focuses on
efficiency and eliminating the waste of excessive inventory
\citep{van2021using}. In JIT systems, accurate Stock Keeping Unit (SKU)
short-term demand forecasts, typically on a daily, weekly or monthly
basis, are crucial as they are the fundamental inputs to drive the
supply chain planning system.Therefore, forecasting performance can
strongly affect a company's financial results, since both under- and
overestimating short-term demand are expensive as they result in supply
shortages and consequently poor customer service, and excess inventories
and product obsolescence, respectively (Sanders and Graman, 2009).In
this section, we provide a brief review of studies on forecasting
harmaceuticals products.In conclusion, the research findings indicated
that the hybrid neural network approach, which combines linear and
nonlinear modeling techniques, outperformed traditional methods like
ARIMA and standalone neural networks in forecasting pharmaceutical
sales. The hybrid approach demonstrated the ability to capture both
linear and nonlinear patterns in sales data accurately, especially in
scenarios with limited historical records for individual drugs.

A study by Khalil Zadeh et al.~developed an intelligent sales prediction
model for Pharmaceutical Distribution Companies (PDCs) by combining
network analysis tools and time series forecasting methods. This
approach addresses the challenge of limited historical sales data for
pharmaceutical products, making traditional forecasting methods
inadequate\citep{zhang2004neural}.The researchers employed Mean Squared
Error (MSE) and Mean Absolute Error (MAE) to evaluate the performance of
various forecasting models, including ARIMA, neural networks, and a
hybrid approach combining both. The hybrid model, inspired by Zhang's
work\citep{zhang2004neural}, used ARIMA for linear components and neural
networks for nonlinear components.Three methods were used: (1) ARIMA,
(2) a hybrid neural network with each drug's past records, and (3) a
hybrid neural network with both each drug's and its group members' past
records. Grouping products and using group members' records
significantly improved forecasting accuracy by addressing the issue of
limited data and leveraging both linear and nonlinear modeling
capabilities.The study concluded that the hybrid approach, incorporating
both individual and group past records, performed better than using
either method alone \citep{khalil2014intelligent}.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{todonotes,mathtools,bm,amsmath,mathpazo}
\mathtoolsset{showonlyrefs}
\setlength{\parindent}{0cm}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{International Journal of Production Research}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting},
  pdfkeywords={Forecasting, Pharmaceutical supply chain, Domain
knowlege, Forecast accuracy, Developing countries},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{The missing puzzle piece: Contextual insights for enhanced
pharmaceutical supply chain forecasting}

@@

\begin{abstract}
Accurate forecasting in pharmaceutical supply chains is critical for
ensuring the continuous availability of essential medicines,
particularly in developing countries where resource constraints and
logistical challenges are more pronounced. Effective demand forecasting
supports procurement and inventory management, which in turn could help
to prevent stockouts, reduce wastage due to overstocking, and enhance
overall healthcare delivery. In such settings, reliable forecasts, that
acknowledge uncertainties, are essential to ensure that limited
resources are used efficiently to meet health needs. Despite
advancements in forecasting, significant gaps remain in effective
forecasting in poor resource countries. Most existing literature on
pharmaceutical supply chain forecasting focuses primarily on point
estimation of consumption while neglecting the inherent uncertainty of
these forecasts. Additionally, these studies often rely solely on
historical consumption data without considering the broader context that
influences consumption. Furthermore, many works lack rigorous
methodological design, transparency, and reproducibility. This study
addresses these gaps by integrating domain-specific knowledge, gathered
through expert interviews and engagement with key members of the
Ethiopian Pharmaceutical Supply Service (EPSS), into forecasting models.
Using a dataset spanning five years (December 2017 to July 2022) from
EPSS, we developed forecasting models that integrate expert-identified
variables such as stock replenishment schedules, fiscal inventory
counts, and disease outbreaks. Evaluation metrics including Mean
Absolute Scaled Error (MASE), Root Mean Squared Scaled Error (RMSSE),
and Continuous Ranked Probability Score (CRPS) are used to report
forecast accuracy. The findings underscore the significance of
contextual data in developing robust forecasting models that are
adaptable to complex, real-world conditions. Our results also highlight
the effectiveness of foundational time series models for forecasting.
These models are particularly appealing for resource-constrained
countries that may lack advanced analytical expertise. To promote
usability, generalizability, and reproducibility, we share the complete
dataset and code in R and Python, and the entire paper is written in
Quarto via a GitHub repository to facilitate these practices.
\end{abstract}

\begin{keyword}
    Forecasting \sep Pharmaceutical supply chain \sep Domain
knowlege \sep Forecast accuracy \sep 
    Developing countries
\end{keyword}
\end{frontmatter}

\section{Introduction}\label{sec-intro}

Universal access to essential medicines is a fundamental goal of
effective healthcare systems, yet ensuring equitable availability
continues to be a critical global challenge
\citep{quick2003essential, world2004annual}. Many countries,
particularly in Africa and Asia, face significant barriers to accessing
these medications, often due to high costs and inefficient supply chain
management \citep{world2004medicines}. A survey conducted in eight
sub-Saharan African countries revealed alarmingly low availability of
essential medicines. The average availability of women's priority
medicines ranged from just 22\% to 40\%, while children's medicines were
only slightly more accessible, with availability ranging from 28\% to
57\% \citep{droti2019poor}. However, drug shortages are not limited to
low-income regions; they also occur in developed areas like the United
States and Europe, where they disrupt healthcare delivery and compromise
the quality of patient care
\citep{fox2003managing, kaakeh2011impact, johnson2011drug, huys2013european, le2011prevalence}.

The consequences of drug shortages extend beyond immediate healthcare
delivery, impacting healthcare costs, treatment quality, and patient
safety \citep{alspach2012drug, kaakeh2011impact, baumer2004national}.
Such shortages can result in treatment delays, suboptimal care due to
the unavailability of preferred therapies, increased reliance on costly
secondary markets, and serious patient outcomes, including
complications, longer hospital stays, and even deaths
\citep{alspach2012drug}. Addressing these challenges requires a
resilient pharmaceutical supply chain that can ensure the timely
availability of medicines. In this context, accurate demand forecasting
becomes critical for effective procurement and inventory management
\citep{subramanian2021effective}.

Demand forecasting in pharmaceutical supply chains is a highly complex
process, shaped by several factors. These include the quality of
available data, the specific position within the supply chain, and
external influences such as market trends, regulatory changes, and
disruptive events such as pandemics, conflicts, flooding
\citep{schneider2010pharmaceutical, hyndman2018forecasting}. Recently,
major donors such as Bill \& Melinda Gates Foundation have invested in
data collection technologies in sub-Saharan Africa, recognizing the
potential of data to provide valuable insights. This investment is
essential and offers significant benefits for improving supply chain
delivery. However, much of the data being collected focuses on
transactional data, such as consumption and sale. Almost all platforms
collecting data tend to overlook the critical contextual factors and
events that influence these transactions---factors that are just as
crucial for accurate forecasting. This presents a major barrier, as
understanding the broader context, including administration procedures,
local policies, disruptions, conflicts, and so on is vital for building
reliable forecasting models. This gap---known as business context and
domain knowledge---requires urgent attention in Pharmaceutical supply
chain forecasting. Without incorporating this knowledge, any forecasting
model is unlikely to produce reliable results. We must prioritize
collecting and integrating this contextual information to improve the
accuracy of demand forecasts in pharmaceutical supply chains. Combining
statistical models with domain insights offers a more refined
forecasting process, as evidenced by research that highlights the value
of judgmental adjustments to enhance forecast accuracy
\citep{fildes2007against, taylor2022artificial, soyiri2013overview}.
Domain knowledge enables analysts to incorporate relevant events, such
as supply chain disruptions, into the forecasting model. This
integration is particularly beneficial in fields like healthcare, where
expert insights can align algorithms with real-world applications
\citep{dash2022review}.

This study aims to bridge that gap by employing advanced machine
learning models combined with domain expertise to address the
distinctive characteristics of Ethiopia's pharmaceutical sector. The
challenges in this sector, such as diverse product categories, limited
data from service delivery points, communication issues, long lead
times, forecast inaccuracies, and complex policies, necessitate tailored
forecasting approaches
\citep{bilal2024challenges, boche2022procurement}. Accurate demand
forecasting in Ethiopia is essential for optimizing resource allocation,
streamlining supply chain operations, and shaping effective healthcare
policies \citep{rostami2022forecasting}.

This research uses consumption data from 33 health commodities, covering
the period from December 2017 to July 2022, provided by the Ethiopian
Pharmaceutical Supply Service (EPSS), to develop models for
six-month-ahead forecasting. Initially, we implement univariate models
that rely solely on consumption data. Following this, we conduct
exploratory data analysis and engage in interviews with EPSS experts to
gather domain knowledge on events influencing consumption patterns.
Through these interviews, the experts identified three key factors
collectively impacting consumption: stock replenishment period, fiscal
calendar, and the seasonal malaria. We then develop forecasting models
that incorporate the data gathered through the interview process and
compare their performance with univariate models that rely solely on
consumption data.

Our primary aims is to assess whether incorporating domain knowledge
improves forecast accuracy. Moreover, we also fill a gap in the
pharmaceutical demand forecasting literature by not only generating
point forecasts but also producing probabilistic forecasting techniques
that capture the range of potential outcomes with associated
probabilities, offering a more comprehensive view of demand uncertainty
than traditional methods. Model performance is evaluated using point
accuracy measure including MASE, RMSSE, and probabilistic measure using
CRPS.

The remainder of this paper is structured as follows: Section 2 reviews
the literature and identifies gaps to position our work; Section 3
outlines the experimental design, including data sources, forecasting
methods, and evaluation criteria; Section 4 presents the results and
analysis; and Section 5 concludes with a summary of findings and future
research directions.

\section{Research background}\label{sec-lit}

Accurate sales forecasting is crucial in the pharmaceutical industry,
where it directly impacts profit maximization, cost minimization, and
the ability to respond to market changes. Forecasting in healthcare not
only influences clinical decisions but also plays a pivotal role in
supply chain management, ensuring that the right drugs are available
when needed. The challenge lies in balancing customer demand with
inventory costs, which is particularly complex in pharmaceuticals due to
factors like short shelf life and quality constraints
\citep{gupta2000mid, makridakis2020forecasting}.

In this review, a range of forecasting methods has been applied, from
traditional statistical models to advanced machine learning techniques,
with many studies using a combination of simple and advanced methods to
enhance accuracy
\citep{nikolopoulos2016forecasting, zhu2021demand, anusha2014demand}.
Some studies, such as those by \citet{newberne2006holt}, and
\citet{restyana2021analysis}, applied traditional forecasting methods
like the Holt-Winters method, Single Moving Average (SMA), and Single
Exponential Smoothing (SES), which are simpler but effective for certain
applications. In contrast, other researchers explored more advanced
approaches, including hybrid models and machine learning techniques, to
address complex forecasting challenges, as seen in the studies by
\citet{siddiqui2022hybrid}, \citet{de2021lead}, \citet{kim2015demand},
\citet{candan2014demand}, and \citet{ribeiro2017improving}.

Early research in pharmaceutical forecasting primarily focused on
traditional statistical methods. These methods, including ARIMA,
exponential smoothing, and moving averages, were widely used due to
their simplicity and effectiveness in relatively stable environments
\citep{zahra2019forecasting}. These models typically assess forecast
accuracy using metrics like Mean Absolute Percentage Error (MAPE), Mean
Squared Error (MSE), and Root Mean Squared Error (RMSE). For example,
\citet{newberne2006holt} applied the Holt-Winters method to forecast
healthcare data, focusing on prescription trends. Their findings
highlighted the method's utility in short-term planning and resource
management, validated through metrics like MAPE, MAD, and MSD.
Similarly, \citet{restyana2021analysis} compared Single Moving Average
(SMA) and Single Exponential Smoothing (SES) in drug demand forecasting,
concluding that SES provided more accurate results, as indicated by
lower MAD and MSE values.

Recognizing the limitations of traditional models, particularly in
capturing the complexities of pharmaceutical demand (e.g., seasonality,
external influences), researchers began developing hybrid models. These
models combine linear methods like ARIMA with nonlinear techniques such
as neural networks, aiming to improve forecasting accuracy by addressing
diverse data patterns. \citet{khalil2014intelligent} introduced a hybrid
model that integrated ARIMA with neural networks. Their study
demonstrated that this approach significantly outperformed traditional
models, especially in scenarios with limited historical data. The hybrid
model's effectiveness was confirmed by improvements in key performance
indicators like MAPE and RMSE.

The pharmaceutical industry's increasing complexity and the availability
of large datasets have led to the adoption of advanced machine learning
techniques. These methods, such as Support Vector Regression (SVR),
Random Forest (RF), and Long Short-Term Memory (LSTM) networks, offer
greater flexibility and predictive power. \citet{van2021using} showed
that incorporating downstream information into machine learning models,
like LASSO and SVR, greatly enhances forecast accuracy. Their study on
multi-echelon supply chains revealed that advanced models, particularly
those integrating external variables, consistently produced lower
forecast errors, as measured by AvgRelRMSE. Similarly,
\citet{rathipriya2023demand} compared shallow neural networks and deep
learning models for drug demand forecasting. The study found that
shallow models outperformed deep learning models for most drug
categories, while ARIMA was more effective for the remaining categories.
This suggests that no single model is universally optimal, and model
selection should be context-specific.

More sophisticated models, such as Neuro-Fuzzy Systems, have emerged as
powerful tools for pharmaceutical forecasting. These systems combine the
learning capabilities of neural networks with the reasoning abilities of
fuzzy logic, offering a balanced approach that incorporates both
empirical data and expert knowledge. Deep learning models, while
powerful, present challenges such as the need for large datasets and
substantial computational resources. Despite these challenges, models
like LSTM have shown potential in capturing long-term dependencies in
time series data, as demonstrated by \citet{sousa2019statistical} in
their study on drug distribution in the Brazilian Public Health System.
\citet{candan2014demand} employed an Adaptive Neuro-Fuzzy Inference
System (ANFIS) to forecast pharmaceutical demand. Their study
highlighted the system's ability to capture complex patterns in the
data, resulting in highly accurate forecasts. The effectiveness of this
approach was further validated through statistical tests, such as paired
T-tests and mean difference analysis, which confirmed its superiority
over traditional methods.

\citet{nguyen2023managing} explore how sentiment analysis of news media
can be utilized to improve demand forecasting for pharmaceutical
products during disruptive events, such as pandemics and scandals. The
authors employed a VARX (Vector Autoregressive with Exogenous Variables)
model to forecast demand volatility. \citet{papanagnou2018coping}
investigate the impact of big data analytics, specifically through text
mining, on improving forecasting accuracy for pharmaceuticals in retail
pharmacies, using HMX pharmacy stores in Nigeria as a case study. They
use multivariate time series analysis technique known as VARX, which
incorporates customer-generated content from sources like Google,
YouTube, and online newspapers to predict demand for medicines.
\citet{fourkiotis2024applying} focus on enhancing pharmaceutical sales
forecasting through the application of both traditional statistical
methods and advanced machine learning techniques. Various forecasting
approaches, including traditional models like Naïve and ARIMA,Facebook
Prophet, and machine learning methods such as XGBoost and LSTM neural
networks are compared. The results demonstrated that machine learning
models, particularly XGBoost, significantly outperformed traditional
methods. \citet{belghith2024new} present a rolling forecasting framework
employing moving average and three exponential smoothing methods
(Brown's, Holt's, and Holt-Winters) implemented using Microsoft Power BI
for enhanced data visualization.

Table~\ref{tbl-literature} provides a summary of key studies in the
literature on forecasting for pharmaceutical products. We identify
several limitations both in the existing research and in practice, which
highlight important gaps that motivate our current study. These gaps are
summarized as follows:
